### 2015-04-13 src/intro/intro.tex
Updated introduction and discussion

One complication is that the magnitude, $e$, of these eye movements not only depend on the translation amplitude, $T$, but also on the fixation depth, $d$. When fixating position $(x, d)$, the eye position at starting position is $e_0 = atan(d/x) \approx d/x$. By subtracting the eye position after the translation, $e_1 = atan(d/(x+T)) \approx d/(x+T)$, we obtain the eye movement amplitude, $e \approx d/T$. In Chapter 4 we explore whether the brain takes this geometry into account and compensates for fixation depth when using eye movements in self-motion perception.

\section{Reflexive uses for the self-motion}
While vection is a valuable cue to self-motion perception, the brain wants to minimize vection because it blurs the image on the fovea, which hampers visual perception. Several reflexive mechanisms attempt to keep stable fixation to avoid image blurring during self-motion. Two such mechanisms are the vestibulo-ocular reflex, the VOR, and the optokinetic reflex, the OKR.

The vestibulo-ocular reflex (VOR) is driven by the short-latency vestibular signal and compensates for both rotation and translation of the head. The rotational and translational part of the VOR have their own dedicated reflex arcs. The translational VOR, or TVOR, is driven by the otolith signal while the rotational VOR, or RVOR, is mainly driven by the semicircular canals. The RVOR operates during head rotation by counter rotating the eyes in the opposite direction to the head. Because its gain is about one, eye velocity is about equal, but of opposite sign, to the head velocity. The TVOR operates during head translation, when the head moves orthogonally to the line of sight. In contrast to the RVOR, the TVOR needs to take fixation depth into account. Simple geometry shows that the ideal TVOR response is inversely proportional to fixation distance, $e_{vor} = atan⁡(d/T)$. As a result, no compensatory eye movements are required when fixating targets at infinity to maintain a stable retinal image.

@@

In many cases the brain does not use a flat prior, but a Gaussian one. For example, in Chapter 2, we use a prior ... head is upright in space.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Add a bit more about optimal integration, perhaps give an example?
%

In this case, the Gaussian prior can be seen as an additional sensory signal that is weighted into the posterior estimate.

@@

A powerful tool in psychophysical research is the two-alternative forced choice, or 2AFC, paradigm. In this paradigm, participants are presented with two stimuli, e.g. two translation distances, and are then forced to make a choice, e.g. on which of the two translations was longer. By systematically varying one of the choice alternatives (i.e., the probe stimulus), the point of subjective equality, or PSE, can be determined. The PSE is the point at which the participant is completely unsure about which stimulus to pick, i.e. gives a random response, because he perceives the two stimuli as being equal. If the PSE does not line up with the actual reference stimulus, there is a bias in the percept, represented by the Greek letter mu ($\mu$).

@@

\subsection{Subjective body tilt}
In the subjective body tilt, or SBT, task the perceived body orientation with respect to a given body tilt angle is probed. Participants are first given a reference angle, e.g. 45 \si{\degree}, and are then rotated to an angle close to the reference angle, e.g. 46 \si{\degree} (see \figref{intro:fig4}A). They then have to indicate whether their current orientation is clockwise or counter-clockwise with respect to the reference angle. By systematically probing rotation angles around the reference angle we can obtain both the bias and uncertainty on the percept of body tilt.

Most participants are able to do this perfectly, regardless of the reference angle (see \figref{intro:fig4}B). Their uncertainty does increase as a function of reference angle though (see \figref{intro:fig4}C).
As the SBT probes body orientation, the somatosensory signal originating from the torso can be used without any reference transformation and thus provides a direct contribution. Other sensory signals such as the vestibular signal can also be used, but only after a reference frame transformation.

@@

    \caption{Subjective body tilt task (SBT); \panelref{A} Graphical representation of the SBT task. \panelref{B} Bias and \panelref{C} uncertainty as a function of roll angle in a typical participant (circles). The dashed line represents the typical pattern.}
    \label{intro:fig4}
\end{figure}

\subsection{Subjective visual vertical}
The subjective visual vertical, or SVV, task is a similar task in which participants have to judge the orientation of a line with respect to gravity. The PSE, that is the angle at which the line is perceived to be aligned with gravity, can be found by presenting lines at different angles and asking the participant whether the line is rotated clockwise or counter-clockwise relative to gravity (see \figref{intro:fig5}A).

When seated straight this task, participants do not make any static errors and are very certain about their responses (see \figref{intro:fig5}B and C). This changes when first rotating the participant before the task. In general, the static error increases with tilt angle, the direction of the error is in the direction of the body mid-line, this effect is known as the Aubert or A-effect. At smaller angles overcompensation occurs and the static error moves in the opposite direction, that is away from the body mid-line. This latter effect is known as the E-effect and could be due to the effects of ocular counter-roll (OCR).

@@

    \caption{Subjective visual vertical task (SVV); \panelref{A} Graphical representation of the SVV task. \panelref{B} Bias and \panelref{C} uncertainty as a function of roll angle in a typical participant (circles). The dashed line represents the typical pattern.}
    \label{intro:fig5}
\end{figure}

@@

Perhaps explain the lateral translation task.

@@

We start with Chapter 2 in which we investigate sensory noise levels of the visual, somatosensory, and vestibular cues to head and body orientation. Because the contributions of these signals cannot be studied in isolation, we adopted an inverse approach where we assumed statistical optimality and attempted to compute the statistical properties of individual sensors. To this end, we fitted an optimal integration model to the behaviour from two psychophyisical tasks. The first gauged head-in-space orientation (SVV) and the second body-in-space orientation (SBT). The resulting estimates of the noise levels in the involved sensory modalities were used to predict the responses of patients with somatosensory and vestibular deficits. These predictions were consistent with previously published deficits for these patient groups, strengthening the idea that human spatial orientation is statistically optimal.

In Chapter 3 we example the dynamic aspects of the integration of vestibular and non-vestibular cues for self-motion perception. We specifically focus on the contributions of eye movements to the perception of whole-body translation. Using a psychophysical task, we first show that translations in which the eyes fixate a body-fixed target are perceived as shorter than those in which the eyes fixate a world-stationary target and thus make pursuit movements. We further demonstrate that this result does not depend on the presence of the fixation point per se, but that the magnitude of unconstrained eye movements in complete darkness directly influences the self-motion percept.

Because fixation distance influences the magnitude of pursuit eye movements, we investigated  whether the brain takes fixation distance into account in Chapter 4. We compared self-motion perception during near and far world-fixed targets. Results suggest that fixation depth is only partially taken into account in the integration of vestibular and non-vestibular information for the calculation of translation distance. It seems that only raw eye movements augment self-motion perception, resulting in a biased percept.

Chapter 5 presents the final experimental study of this thesis, where we investigate how self-motion signals are used to update the location of previously seen objects across whole-body translation. We show that the errors made during spatial updating do not only depend on the location of the previously seen object, but also on the location of gaze. This is consistent with spatial updating in a gaze-centred reference frame. We further show that the underestimation of self-motion is a possible cause of the errors found.

>>>>

One complication is that the magnitude, $e$, of these eye movements not only depend on the translation amplitude, $T$, but also on the fixation depth, $d$. Oculomotor cues to depth include binocular vergence angle and accommodation, both most robust for less fixations than a meter or so away(ref). When fixating position $(x, d)$, the eye position at starting position is $e_0 = atan(d/x) \approx d/x$. By subtracting the eye position after the translation, $e_1 = atan(d/(x+T)) \approx d/(x+T)$, we obtain the eye movement amplitude, $e \approx d/T$. In Chapter 4 we explore whether the brain takes this geometry into account and compensates for fixation depth when using eye movements in self-motion perception. Fixation depth plays also a crucial role in chapter 5, in which the the effects of eye and head movements on spatial updating are investigated

\section{Reflexive uses for the self-motion}
While vection is a valuable cue to self-motion perception, the brain wants to minimize vection because it blurs the image on the fovea, which hampers visual perception. Several reflexive mechanisms attempt to keep stable fixation to avoid image blurring during self-motion. Two such mechanisms are the vestibulo-ocular reflex, the VOR \cite{goldberg2012}, and the optokinetic reflex, the OKR \cite{purves2012}.

The vestibulo-ocular reflex (VOR) is driven by the short-latency vestibular signal and compensates for both rotation and translation of the head. The rotational and translational part of the VOR have their own dedicated reflex arcs. The translational VOR, or TVOR, is driven by the otolith signal while the rotational VOR, or RVOR, is mainly driven by the semicircular canals. The RVOR operates during head rotation by counter rotating the eyes in the opposite direction to the head. Because its gain is about one, eye velocity is about equal, but of opposite sign, to the head velocity \cite{goldberg2012}. The TVOR operates during head translation, when the head moves orthogonally to the line of sight. In contrast to the RVOR, the TVOR needs to take fixation depth into account. Simple geometry shows that the ideal TVOR response is inversely proportional to fixation distance, $e_{vor} = atan⁡(d/T)$. As a result, no compensatory eye movements are required when fixating targets at infinity to maintain a stable retinal image.

@@

In many cases the brain does not use a flat prior, but a Gaussian one. For example, in Chapter 2, we use a prior ... that encodes the assumption that the head is mostly upright in space. In this case, the Gaussian prior can be seen as an additional sensory signal that is weighted into the posterior estimate.

Statistically optimal integration has been shown in many situations, for example in sensorimotor learning \cite{kording2004}, spatial localization \cite{battaglia2003} and updating \cite{vaziri2006}.

@@

A powerful tool in psychophysical research is the two-alternative forced choice (2AFC) paradigm. In this paradigm, participants are presented with two stimuli, e.g. two translation distances, and are then forced to make a choice, e.g. on which of the two translations was longer. By systematically varying one of the choice alternatives (i.e., the probe stimulus), the point of subjective equality, or PSE, can be determined. The PSE is the point at which the participant is completely unsure about which stimulus to pick, i.e. gives a random response, because he perceives the two stimuli as being equal. If the PSE does not line up with the actual reference stimulus, there is a bias in the percept, represented by the Greek letter mu ($\mu$).

@@

\subsection{Subjective body tilt}
In the subjective body tilt, or SBT, task the perceived body orientation with respect to a given body tilt angle is probed. Participants are first given a reference angle, e.g. 45 \si{\degree}, and are then rotated to an angle close to the reference angle, e.g. 46 \si{\degree} (see \panelref{intro:fig4}{A}). They then have to indicate whether their current orientation is clockwise or counter-clockwise with respect to the reference angle. By systematically probing rotation angles around the reference angle we can obtain both the bias and uncertainty on the percept of body tilt.

Most participants are able to do this perfectly, regardless of the reference angle (see \panelref{intro:fig4}{B}). Their uncertainty has been shown to increase as a function of reference angle though (see \panelref{intro:fig4}{C}).

As the SBT probes body orientation, the somatosensory signal originating from the torso can be used without any reference transformation and thus provides a direct contribution. Other sensory signals such as the vestibular signal can also be used, but only after a reference frame transformation.

@@

    \caption{Subjective body tilt task (SBT); \panel{A} Graphical representation of the SBT task. \panel{B} Bias and \panel{C} uncertainty as a function of roll angle in a typical participant (circles). The dashed line represents the typical pattern.}
    \label{intro:fig4}
\end{figure}

\subsection{Subjective visual vertical}
The subjective visual vertical, or SVV, task is a similar task in which participants have to judge the orientation of a line with respect to gravity. The PSE, that is the angle at which the line is perceived to be aligned with gravity, can be found by presenting lines at different angles and asking the participant whether the line is rotated clockwise or counter-clockwise relative to gravity (see \panelref{intro:fig5}{A}).

When seated straight this task, participants do not make any static errors and are very certain about their responses (see \panelref{intro:fig5}{B} and \hyperref[intro:fig5]{C}). This changes when tilting the participant before the task. In general, the static error increases with tilt angle. When the direction of the error is in the direction of the body mid-line, this effect is known as the Aubert or A-effect. At smaller angles overcompensation occurs and the static error is in the opposite direction, that is away from the body mid-line. This latter effect is known as the E-effect and could be due to the effects of ocular counter-roll (OCR).

@@

    \caption{Subjective visual vertical task (SVV); \panel{A} Graphical representation of the SVV task. \panel{B} Bias and \panel{C} uncertainty as a function of roll angle in a typical participant (circles). The dashed line represents the typical pattern.}
    \label{intro:fig5}
\end{figure}

@@

In the lateral translation task participants had to judge which of two subsequent lateral translations was longest in magnitude. By manipulating one of these translations while keeping the other constant, the effects of these manipulations on self-motion perception can be studied. Because we ask participants to report on the difference between two translations (i.e. longer versus shorter), only relative effects can be assessed.  We used the lateral translation task in \chapter{p3,p4} to assess the effects of fixation type and distance on self-motion perception.

@@

We start with \chapter{p1} in which we investigate sensory noise levels of the visual, somatosensory, and vestibular cues to head and body orientation. Because the contributions of these signals cannot be studied in isolation, we adopted an inverse approach where we assumed statistical optimality and attempted to compute the statistical properties of individual sensors. To this end, we fitted an optimal integration model to the behaviour from two psychophyisical tasks. The first gauged head-in-space orientation (SVV) and the second body-in-space orientation (SBT). The resulting estimates of the noise levels in the involved sensory modalities were used to predict the responses of patients with somatosensory and vestibular deficits. These predictions were consistent with previously published deficits in these patient groups, strengthening the idea that human spatial orientation is statistically optimal.

In \chapter{p3} we examine the dynamic aspects of the integration of vestibular and non-vestibular cues for self-motion perception. We specifically focus on the contributions of eye movements to the perception of whole-body translation. Using a psychophysical task, we first show that translations in which the eyes fixate a body-fixed target are perceived as shorter than those in which the eyes fixate a world-stationary target and thus make pursuit movements. We further demonstrate that this result does not depend on the presence of the fixation point per se, but that the magnitude of unconstrained eye movements in complete darkness directly influences the self-motion percept.

Because fixation distance influences the magnitude of pursuit eye movements, we investigated  whether the brain takes fixation distance into account in \chapter{p4}. We compared self-motion perception during gaze fixation on near and far world-fixed targets. Results suggest that fixation depth is only partially taken into account in the integration of vestibular and non-vestibular information for the calculation of translation distance. It seems that only raw eye movements augment self-motion perception, resulting in a biased percept.

An accurate internal estimate of self-motion perception is not only crucial for navigation, but can also be used to updated memorized locations of previously seen objects. This is process in known as spatial updating. \chapter{p2} presents the final experimental study of this thesis, where we investigate how self-motion signals are used in spatial updating. We show that the errors made during spatial updating do not only depend on the location of the previously seen object, but also on the location of gaze. This is consistent with spatial updating in a gaze-centred reference frame. We further show that the underestimation of self-motion is a possible cause of the errors found.

### 2015-04-13 src/misc/s2_discussion.tex
Updated introduction and discussion

Navigating through the environment evokes complex changes of visual, auditory, vestibular, tactile and
motor inputs to the brain. Yet, despite this motion, we perceive the world as a stable reality, maintain an integrated sense of where we are, how we are oriented, and are able to track and act rather effortlessly upon surrounding objects.  How the ability comes about it one of the key questions in neuroscience and also studied in this research project. The objective was to build computational models and perform thorough psychometric testing in order to examine the physical and biological contrast that are placed on the interaction between the vestibular and other sensory systems for spatial orientation and self-motion perception. This resulted in the following contributions to the field:

\begin{enumerate}
\item Statistical optimality can account for the way body somatosensory, neck proprioceptive and vestibular signals are integrated in spatial orientation perception. While these sensors cannot be examined in isolation, by using optimality as a starting point, their noise properties can be determined and linked to clinical deficits that are seen in particular patient groups.
\item Oculomotor signals influence self-motion perception, even in the absence of optic flow or other visual stimulation, and even when in conflict with vestibular information. 
\item This oculomotor signals should be regarded as rudimentary cue to self-motion perception; it is not veridically scaled by fixation distance in the perception of body translation. 
\item Self-motion signals interact with the dynamic perception of external world.  Errors that arise in this process suggest the use of a gaze-centered reference frames in the underlying computations. 
\end{enumerate}

@@

\section{Multisensory processing in spatial orientation}
Many studies have shown that the brain tends to combine noisy sensory signals in a statistically optimal way. This is normally done by showing that the noise levels of each separate modality can predict performance in combined conditions. In spatial orientation it is difficult to measure sensory signals in isolation as multisensory integration always takes place. In Chapter 2 we therefore used a statistically optimal integration model as a starting point, and attempted to account for the perceptual differences found when probing head-in-space versus body-in-space orientation. Using a psychometric approach, we tested both the perception of body tilt (subjective body tilt; SBT) as well as the perception of visual vertical (subjective visual vertical; SVV) in seven participants. Our model was able to explain the results from both tests. Because one of sensory modalities used, the noise in the neck proprioceptors, can be measured in relative isolation, we used it to independently confirm the predicted head-on-body noise found by the model. We further validated our model by showing that predictions made by our model are consistent with previously published deficits in vestibular and somatosensory patients. We conclude that Bayesian computations can account for the typical differences in spatial orientation judgments associated with different task requirements.

\section{Eye movements influence self-motion perception}
Eye movement typically accompany self-motion in order to minimize retinal slip and maximize dynamic visual acuity. In Chapter 3 we investigate whether these eye movements also have a reversed role, serving as a cue for self-motion perception. To address this question, we asked participants to compare perceived translation distance in two successive, passive, lateral whole-body translations. Eye movements during these translations were either world-stationary or body-stationary. Results show that translations were perceived shorter with body-fixed gaze compared to world-fixed gaze, indicating that eye movements indeed influence self-motion perception. Using a linear model, we estimated the relative contribution of the vestibular versus the eye movement signal, and found that eye movement signals contribute approximately 25 percent of the perceived motion. We independently validated the model by successfully predicting the effects of eye movements on self-motion perception during trails in which the eye movements were unconstrained. This further shows that eye movement signals influence self-motion perception, even in the absence of visual stimulation, and even when oculomotor and vestibular estimates are in conflict, e.g. during body-fixed gaze. We hypothesize that adverse consequences of this seemingly inflexible arrangement are minimal under natural conditions because eye movements and self-motion are highly correlated, and because eye movements are most often accompanied by veridical optic flow cues to self-motion.

\section{Partial compensation for fixation depth in self-motion perception}
The amplitude of these eye movements depend on the depth of fixation, when fixating  faraway they are smaller than when fixating nearby in the world. If these eye movements are used to augment self-motion perception, the effect of fixation depth has to be taken into account in order for self-motion perception to be veridical. In Chapter 4 we investigate whether the brain indeed compensates and takes fixation depth into account. Participants had to judge self-motion during different eye movement conditions in the absence of full-field optic flow. In a 2-AFC task, similar to the one used in Chapter 3, participants indicated whether the second of two successive passive lateral whole-body translations was longer or shorter than the first. During each translation, participants fixated either a nearby or far away target, which was either body- or world-stationary. Results show that the perceived translations were shorter for nearby world-fixed gaze compared to far away world-fixed gaze, indicating that eye movements are not properly scaled in self-motion perception. Together with the observation that self-motion perception is not affected by the depth of a body stationary fixation target, we conclude that eye movements are merely a rudimentary cue to self-motion, with a compensation for fixation depth that is partial at best.

\section{Gaze-dependent effects in spatial updating}
The brain uses these self-motion signals to update ego-centric spatial representations of the environment. In Chapter 5 we investigate how, and to which extent, the brain integrates these various self-motion signals for the spatial update. Participants were oscillated sideways while keeping gaze fixed on a stationary target. When the motion direction changed, a reference target was shown either in front or behind the fixation point. Half a cycle later, at the next reversal, we tested updating of this reference location by asking participants to judge whether a briefly flashed probe was shown to the left or right of the memorized target. Results show that both the direction and magnitude of the bias in spatial updating depends on the location of the object being updated with respect to gaze, implying that a gaze-centered reference frame is involved. We further show that these biases can be caused by an underestimated of translation amplitude, a bias of visually perceived objects towards the fovea, or by a combination of both.

@@

To be done.

%Pieter's suggestions: 10-line summary, why optimality?, questions that remain, where do we go in the future (neurophysiology, neurostimulation, modeling, ...).

%[Could it be that the fixation target is a world-fixed reference? Add something about that here or later.]

%%%%%%%%%%%%%
% Integrated model

%Could check Anna/Alexandra results again -> updating + self-motion effect.
%Reaching poster I've made for SfN

%%%%%%%%%%%%%%
% Relation to other studies

%Based on our updating and self-motion perception studies -> is it possible to make predictions for other studies?
%What would we predict van Pelt would do for instance?

%%%%%%%%%%%%%%
% Conclusion

>>>>

Navigating through the environment evokes complex changes of visual, auditory, vestibular, tactile and
motor inputs to the brain. Yet, despite these motion induced changes of input, we perceive the world as a stable reality, maintain an integrated sense of where we are, how we are oriented, and are able to track and act rather effortlessly upon surrounding objects. How this ability comes about is also the topic of this thesis. The objective of the research project described in this thesis was to build computational models and perform thorough psychometric testing in order to unravel the physical and biological constraints on the interaction between the vestibular and other sensory systems for spatial orientation and self-motion perception. This resulted in the following contributions to the field:

\begin{description}
\item[\chapter{p1}] Statistical optimality can account for the way body somatosensory, neck proprioceptive and vestibular signals are integrated in spatial orientation perception. While these sensors cannot be examined in isolation, by using optimality as a starting point, their noise properties can be determined and linked to clinical deficits that are seen in particular patient groups.
\item[\chapter{p3}] Oculomotor signals influence self-motion perception, even in the absence of optic flow or other visual stimulation, and even when in conflict with vestibular information. 
\item[\chapter{p4}] This oculomotor signals should be regarded as rudimentary cue to self-motion perception; it is not veridically scaled by fixation distance in the perception of body translation. 
\item[\chapter{p2}] Self-motion signals interact with the dynamic perception of external world.  Errors that arise in this process suggest the use of a gaze-centered reference frames in the underlying computations. 
\end{description}

@@

\section{Multisensory processing in spatial orientation}
Many studies have shown that the brain combines noisy sensory signals in a statistically optimal way. This is normally done by showing that the noise levels of each separate modality can predict performance in combined conditions. In spatial orientation it is difficult to measure sensory signals in isolation as one cannot switch off the vestibular sense when measuring somatosensory contributions. In \chapter{p1} we therefore used a statistically optimal integration model as a starting point, and attempted to account for the perceptual differences found when probing head-in-space versus body-in-space orientation. Using a psychometric approach, we tested both the perception of body tilt (subjective body tilt; SBT) as well as the perception of visual vertical (subjective visual vertical; SVV) in seven participants. Because both the SVV and SBT make use of the same sensory signals, we were able to fit a 7-parameter probabilistic model to the response data.  One  of the estimated parameters represented the noise of the neck proprioceptors. This allowed us to independently confirm that the derived values for neck noise matched those that were measured in isolation. We further validated our model by showing that predictions made by our model are consistent with previously published deficits in vestibular and somatosensory patients. We conclude that Bayesian computations can account for the typical differences in spatial orientation judgments associated with different task requirements. In a follow-up to this work, this approach was recently applied to a patient population with complete  vestibular loss \cite{alberts2015}. Performance in those patients was similar to that of controls, suggesting that the sensory weights had shifted from the vestibular to the somatic sensors.

\section{Eye movements influence self-motion perception}
Eye movement typically accompany self-motion in order to minimize retinal slip and maximize dynamic visual acuity. In \chapter{p3} we investigated whether these eye movements also have a reversed role, by serving as a cue for self-motion perception. To address this question, we asked participants to compare perceived translation distances from two successive, passive, lateral whole-body translations. Eye movements during these translations were either world-stationary or body-stationary. Results show that translations were perceived shorter with body-fixed gaze compared to world-fixed gaze, indicating that eye movements indeed influence self-motion perception. Using a linear model, we estimated the relative contribution of the vestibular versus the eye movement based displacement signal: the eye movement based displacement signal  contributes approximately 25 percent to the perceived motion. We independently validated the model by successfully predicting the effects of eye movements on self-motion perception during trials in which the eye movements were unconstrained. This further shows that eye movement signals influence self-motion perception, even in the absence of visual stimulation, and even when oculomotor and vestibular estimates are in conflict, e.g. during body-fixed gaze. We hypothesize that adverse consequences of this seemingly inflexible arrangement are minimal under natural conditions because eye movements and self-motion are highly correlated, and because eye movements are most often accompanied by veridical optic flow cues to self-motion.

\section{Partial compensation for fixation depth in self-motion perception}
If eye movements are used in the estimation of self-motion magnitude, the brain should also take the accompanying fixation depth into account for a veridical translation estimate. The reason is that the amplitude of these eye movements, for the same physical translation, depends on the depth of fixation: when fixating faraway they are smaller than when fixating nearby in the world.  In \chapter{p4} we investigated whether the brain indeed takes fixation depth into account when using eye movements to augment self-motion perception. Participants had to judge self-motion during different eye movement conditions in the absence of full-field optic flow. In a 2-AFC task, similar to the one used in \chapter{p3}, participants indicated whether the second of two successive passive lateral whole-body translations was longer or shorter than the first. During each translation, participants fixated either a nearby or far away target, which was either body- or world-stationary. Results show that the perceived translations were shorter for nearby world-fixed gaze compared to faraway world-fixed gaze, indicating that eye movements are not properly scaled in self-motion perception. Together with the observation that self-motion perception is not affected by the depth of a body stationary fixation target, we conclude that eye movements are merely a rudimentary cue to self-motion, with a compensation for fixation depth that is partial at best.

\section{Gaze-dependent effects in spatial updating}
The brain also needs self-motion signals to update ego-centric spatial representations of the environment. In \chapter{p2} we investigate how, and to which extent, the brain integrates the various self-motion signals for the spatial update. Participants were oscillated sideways while keeping gaze fixed on a stationary target. When the motion direction changed, a reference target was shown either in front or behind the fixation point. Half a cycle later, at the next reversal, we tested updating of this reference location by asking participants to judge whether a briefly flashed probe was shown to the left or right of the memorized target. Results show that both the direction and magnitude of the bias in spatial updating depends on the location of the object being updated with respect to gaze, implying that a gaze-centered reference frame is involved. We further show that these biases can be caused by an underestimated of translation amplitude, a bias of visually perceived objects towards the fovea, or by a combination of both.

@@

All experimental chapters in this thesis report on subjects’ spatial perception. In \chapter{p1} we use an optimal statistical integration model of all the contributing sensory systems to explain both the bias and precision of these percepts. In these models the contributions of the senses are weighted according to their uncertainty. In the chapters on lateral displacement perception (\chapter{p3,p4}) we also use weightings of the vestibular and eye-movement based estimates of displacement. However, these weights are only based on the observed biases in self-motion perception. Ideally one would use the same approach as in \chapter{p1} to model and analyze these self-motion perception data. We have made initial efforts to arrive at such a model, but there appear at least two major challenges. First, the geometry of the self-motion perception experiments makes that the straightforward Gaussian distributions from the spatial orientation model become less well defined, skewed distributions and we would have use particle filter models to run our simulations. Second, those models contain many free parameters for the sensory modalities and priors. With the current set of experiments we do not have enough data to fit these parameters in a consistent manner. Future research should be done to derive a complete optimal integration account of self-motion perception.

In \chapter{p2} we show that eye movements elicited by fixating a world-stationary target are taken into account when updating the remembered location of previously seen targets.  If the underlying signals used for self-motion perception as found in \chapter{p3,p4} are also used for spatial updating, the type of fixation (world- versus body-stationary) should have an effect on the observed biases in spatial updating as well. Indeed, results from a pilot experiment in the early stage of this thesis suggest that fixation type influences spatial updating performance under translation \cite{Clemens2010}. Participants had to reach towards remembered targets after an intervening translation. Preliminary results suggest that reaching errors are large when fixating a body-stationary target during translation, while no such errors exist while fixating world-stationary targets \cite{clemens2010}, suggesting that indeed the eye-movement based self-motion cues are also used in spatial updating.

In conclusion, I hope to have made new advances in the understanding of the mechanisms for spatial orientation and self-motion perception. Of course, there are many remaining questions for further study. These studies should primarily focus on the computational and theoretical mechanisms, and how they are implemented by the neural circuitry. These studies should not only address the computational and theoretical mechanisms but also on the neural implementation and pathways that can be found in the brain.
